% -*- root: ../thesis.tex -*-
\pdfbookmark[1]{Acknowledgments}{acknowledgments}



\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Acknowledgments}
\thispagestyle{empty}

I thank all the persons that supported me during my studies and this dissertation.

\noindent My special thanks to my supervisor Prof. Dr. Ralf B. Schäfer for his support throughout the last six years. 
I am thankful for his openness to my ideas and the opportunities given to follow them, 
for organizing funding throughout this dissertation, 
for pushing me to sound scientific writing and critical reading, 
for challenging discussions, not only on statistical eco(-toxico)logy but also outside of the subject area.

\noindent Many thanks to Prof. Dr. Ralf Schulz for examining this thesis and his influence on me during my undergrad studies.
Moreover, I thank Prof. Dr. Engelbert Niehaus for chairing the dissertation committee.
%! Auch dem Drittgutachter und Niehaus danken.

\noindent Without the continuous support of my parents, Anca and Helmut, this thesis would not have been possible - Thank you!

\noindent I am grateful to my colleagues, students and other people for asking me tough statistical questions. 
These questions from a broad range of fields and finding solutions to them widened my expertise in the field.

\noindent Special thanks to Phillip Uhl, Gunnar Oehmichen and Michael Burstert for the discussions during coffee breaks and proofreading this thesis.

\noindent I thank all collaborators of projects involved in this thesis, but also past projects. All of you provided help, critical comments and enlightening discussions on my work.

\noindent I tried to make this thesis as open and reproducible as possible. I would thank the people developing, maintaining and bug fixing the open software I used throughout this thesis. I thank GitHub for providing me a discount the last three years and a platform for collaboration and version control that was crucial for big parts of this thesis. 

\noindent Lastly, I cannot thank enough my girlfriend, Anja Loescher, for getting through this stressful time with me.
Despite a hard time with her own profession, she always supported, encouraged and loved me. 


\endgroup 
